---
output: rmdwiki::mediawiki_document
---

```{r library_setup, include=FALSE}
library(knitr)
library(curl)
library(dplyr)
library(jsonlite)
library(purrr)
library(DT)
library(data.table)
library(formattable)
library(kableExtra)
```
# Principal Cities

Principal Cities are defined by the Office of Budget and Management (OMB) based on American Community Survey (ACS) 5yr data. OMB releases updated lists via Federal Bulletins at various times during the calendar year. Due to the release date of ACS data, the data released in each bulletin typically lags the release year by approximately 2 years. Since 2010, there have been five (5) bulletins releasing updates to the list of Principal Cities in the United States. The bulletins and their respective release dates (along with ACS data used) are:

  * OMB Bulletin No. 18-04 [released 2018-09-14 using ACS 2015 5yr data](https://www.whitehouse.gov/wp-content/uploads/2018/09/Bulletin-18-04.pdf)
  * OMB Bulletin No. 17-01 [released 2017-08-15 using ACS 2014 5yr data](https://www.whitehouse.gov/sites/whitehouse.gov/files/omb/bulletins/2017/b-17-01.pdf)
  * OMB Bulletin No. 15-01 [released 2015-07-15 using ACS 2013 5yr data](https://www.whitehouse.gov/sites/whitehouse.gov/files/omb/bulletins/2015/15-01.pdf)
  * OMB Bulletin No. 13-01 [released 2013-02-28 using Census and ACS 2010 data](https://www.whitehouse.gov/sites/whitehouse.gov/files/omb/bulletins/2013/b13-01.pdf)
  * OMB Bulletin No. 10-02 [released 2009-12-01 using ACS 2007 data](https://www.whitehouse.gov/sites/whitehouse.gov/files/omb/bulletins/2010/b10-02.pdf)

From 2010-2017, there were eleven (11) Principal Cities designated in the two (2) Metropolitan Statistical Areas (MSA) in the four counties that comprise PSRC. In 2018, a 12th Principal City was added to the list for the region. The twelve (12) currently designated Principal Cities in the PSRC region are:

  1. Auburn (added in 2010)
  2. Bellevue
  3. Bremerton
  4. Everett
  5. Kent
  6. Lakewood
  7. Port Orchard (added in 2018)
  8. Redmond
  9. Renton
  10. Seattle
  11. Silverdale (a Census Designated Place)
  12. Tacoma

This document pulls together data from the Census API for all available ACS 5yr data releases since 2011 and compares them to the Principal City criteria to show how the data and designations vary by year.


```{r data_setup_and_calcs, include=FALSE}

api_key <- '6d9263105b3ca3213e093323b4ece211ab49d4e5'
acs_years <- list('2018','2017', '2016', '2015', '2014')

current_principal_cities <- list("Auburn"="03180",
                                 "Bellevue"="05210",
                                 "Bremerton"="07695",
                                 "Everett"="22640",
                                 "Kent"="35415",
                                 "Lakewood"="38038",
                                 "Port Orchard" ="55785",
                                 "Redmond"="57535",
                                 "Renton"="57745",
                                 "Seattle"="63000",
                                 "Silverdale"="64365",
                                 "Tacoma"="70000")

current_principal_cities_msa <- list("Auburn"="Seattle-Tacoma-Bellevue",
                                 "Bellevue"="Seattle-Tacoma-Bellevue",
                                 "Bremerton"="Bremerton-Silverdale-PortOrchard",
                                 "Everett"="Seattle-Tacoma-Bellevue",
                                 "Kent"="Seattle-Tacoma-Bellevue",
                                 "Lakewood"="Seattle-Tacoma-Bellevue",
                                 "Port Orchard" ="Bremerton-Silverdale-PortOrchard",
                                 "Redmond"="Seattle-Tacoma-Bellevue",
                                 "Renton"="Seattle-Tacoma-Bellevue",
                                 "Seattle"="Seattle-Tacoma-Bellevue",
                                 "Silverdale"="Bremerton-Silverdale-PortOrchard",
                                 "Tacoma"="Seattle-Tacoma-Bellevue")

principal_city_tables <- list("Total Population"="DP05_0001E",
                              "Population 16+"="DP03_0001E",
                              "Resident Workers 16+"="DP03_0018E",
                              "All Workers 16+"="B08604_001E")

largest_principal_cities <- list("Bremerton-Silverdale-PortOrchard"="07695","Seattle-Tacoma-Bellevue"="63000")

pull_api_result <- function(data_year, current_variable, current_geography, user_key) {
  
  # URL for Census API
  if (substr(current_variable, start = 1, stop = 2) == "DP") {
    census_api_call <- paste0('https://api.census.gov/data/', data_year, '/acs/acs5/profile?get=', current_variable, '&for=place:',current_geography, '&in=state:53&key=',user_key)
  } else {
    census_api_call <- paste0('https://api.census.gov/data/', data_year, '/acs/acs5/?get=', current_variable, '&for=place:',current_geography, '&in=state:53&key=',user_key)
  }
  
  # Download data and process the json output into an intger result and return value
  current_data <- curl_fetch_memory(census_api_call)
  interim <- prettify(rawToChar(current_data$content))
  current_result <- fromJSON(interim)
  result <- as.integer(current_result[[2]]) 
  
  return(result)
}

```

# Principal Cities as of OMB Bulletin No. 18-04
As noted above, there are twelve (12) Principal Cities designated in the PSRC region as of September 2018. The criteria for principal city designation have changed over time but the criteria have remained the same since the Federal Register from 2010-06-28. The four (4) criteria that can be used in the designation of a Principal City are:

  A) The largest incorporated place with a Census Population of at least 10,000 in the Core-Based Statistical Areas (CBSA) or, if no incorporated place of at least 10,000 population is present in the CBSA, the largest incorporated place or census designated place in the CBSA.

  B) Any additional incorporated place or census designated place with a population of at least 250,000 or in which 100,000 or more people work.

  C) Any additional incorporated place or census designated place with a population of at least 50,000 but less than 250,000 and in which the number of workers working in the place meets or exceeds the number of workers living in the place.

  D) Any additional incorporated place or census designated place with a population of at least 10,000 but less than 50,000 and at least one-third (1/3) the population size of the largest place and in which the number of workers working in the place meets or exceeds the number of workers living in the place. 

The tabs below summarize the data by the current 12 designated principal cities and show the criteria (A-D) that the city meets with that specific Census data. A "-" indicates that that city did not meet any of the criteria for a Principal City using that years ACS 5yr data.



```{r current-cities, echo = FALSE, include = FALSE, results = 'asis'}
out <- NULL

for (i in acs_years) {
  latest_acs_year <- i
  working_cities <- current_principal_cities
  working_cities_msa <- current_principal_cities_msa
  list_length = 15
  out <- c(out, knitr::knit_child("data-by-year-child-kable.Rmd"))
}
```

`r paste(out, collapse="\n")`
